Improved Particle Approximations \\ to the Joint Smoothing Distribution \\ Using Markov Chain Monte Carlo

### 2012-01-25 Paper/draft3_MCMCsmoother.tex
Complete first draft. Minor changes to plotting.

\title{Faster particle smoothing using Markov chain Monte Carlo sampling}

@@

\begin{abstract}
The abstract goes here.
\end{abstract}

\begin{IEEEkeywords}

\end{IEEEkeywords}

@@

Each particle is an entire realisation of the state process, with $K$ values corresponding to the state at each time step. Methods for generating these samples have been described in \cite{Kitagawa1996,Godsill2004,Briers2010}. In this paper new methods are presented for drawing samples from the smoothing distribution, using Markov chain Monte Carlo (MCMC). These novel procedures have computational advantages and greater flexibility over previous methods.

In section~\ref{sec:basics}, we review briefly review the basic particle filter and existing joint smoother algorithms. The new MCMC-based smoothing methods are presented in sections~\ref{sec:mcmc_smoother} and~\ref{sec:new_state_smoother} with supporting simulations in section~\ref{sec:simulations}.

@@

The particle filter is summarised in algorithm~\ref{alg:PF}.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
  	\FOR{$k = 1 \dots K$}
  		\FOR{$i = 1 \dots N_F$}
  			\STATE Sample $\{ x_{k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
  			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
  			\STATE Discard $x_{k-1}^{(i)}$.
  		\ENDFOR
  	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
  	\ENDFOR
  \end{algorithmic}
  \caption{Particle filter algorithm}
  \label{alg:PF}
\end{algorithm}

@@

The filter-smoother is summarised in algorithm~\ref{alg:FS}.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
  	\FOR{$k = 1 \dots K$}
  		\FOR{$i = 1 \dots N_F$}
  			\STATE Sample $\{ x_{1:k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
  			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
  		\ENDFOR
  	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
  	\ENDFOR
  \end{algorithmic}
  \caption{Filter-smoother algorithm}
  \label{alg:FS}
\end{algorithm}

An improved particle smoothing algorithm was presented in \cite{Godsill2004}. This works by constructing new particles from the smoothing distribution by resampling from the filtering approximations. This has the limitation that the particle support for the smoother is limited to that of the filter. The sampling procedure exploits the following factorisation of the joint distribution.

@@

Particles are generated by sampling backwards in time from the factors of this expansion. At the $k^{th}$ step, a sample $\tilde{x}_{k+1:K} \sim p(x_{k+1:K}|y_{1:K})$ will already have been drawn. By sampling $x_k \sim p(x_k|\tilde{x}_{k+1}, y_{1:K})$ and appending $x_k$ to $\tilde{x}_{k+1:K}$, the state sequence is sequentially extended back in time. The recursion is initialised with $x_K \sim p(x_K|y_{1:K})$, i.e. a sample from the final filtering approximation.

@@

The procedure may be repeated to produce as many samples from the joint smoothing distribution as required. This backward-resampling smoother is summarised in algorithm~\ref{alg:DBRS}.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Run a particle filter to approximate $p(x_k|y_{1:k})$ for each $k$.
		\FOR{$i = 1 \dots N_S$}
			\STATE Sample $\tilde{x}_{K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{K}^{(j)}}(x_{K})$.
			\FOR{$k = K-1 \dots 1$}
				\FOR{$j = 1 \dots N_F$}
					\STATE Calculate weight $\tilde{w}_k^{(j)} = w_k^{(j)} p(\tilde{x}_{k+1}|x_k^{(j)})$.
				\ENDFOR
				\STATE Sample $\tilde{x}_{k}^{(i)} \sim \sum_j \tilde{w}_k^{(j)} \delta_{x_{k}^{(j)}}(x_{k})$.
			\ENDFOR
		\ENDFOR
  \end{algorithmic}
  \caption{Direct backward-resampling smoother algorithm}
  \label{alg:DBRS}
\end{algorithm}

A method for drawing samples from the joint smoothing distribution, $p(x_{1:K}|y_{1:K})$, is included in \cite{Briers2010}. The principal topic of this paper is a two-filter smoother for estimation of the marginal smoothing distributions, $p(x_{k}|y_{1:K})$, by combination of the results from one forward and one backward particle filter. Particles from the joint distribution may then be generated by resampling from either the forward or backward filters in the manner of algorithm~\ref{alg:DBRS}.

@@

In section~\ref{sec:mcmc_smoother}, we introduce an MCMC approach for drawing samples from the joint smoothing distribution with an improved computational complexity over the direct backward-resampling smoother. In section~\ref{sec:new_state_smoother}, we combine this procedure with the new-proposal method used in \cite{Fearnhead2010} to generate joint smoothing distribution particles with full support.

@@

The bottleneck of the backward-resampling smoother (algorithm~\ref{alg:DBRS}) is the calculation of the sampling weights (equation~\ref{eq:DBRS_weights}). For each smoother particle and for each time step, a weight must be calculated for each of the $N_F$ filter particles, in order to draw a sample from the conditional distribution $\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K})$. The innovation here is to use Markov chain Monte Carlo (MCMC) to produce these samples instead of sampling directly.

@@

The complete MCMC backward-resampling smoother is summarised in algorithm~\ref{alg:MCMC-BRS}.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
		\FOR{$i = 1 \dots N_S$}
			\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
			\FOR{$k = K-1 \dots 1$}
				\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
				\FOR{$m = 1 \dots M$}
					\STATE Propose a new state, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
					\STATE With probability $\alpha = \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }$, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
				\ENDFOR
				\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{k}^{(i)(M)}$
			\ENDFOR
		\ENDFOR
  \end{algorithmic}
  \caption{MCMC backward-resampling smoother algorithm}
  \label{alg:MCMC-BRS}
\end{algorithm}

@@

The computational complexity of the MCMC-based algorithm is $\mathcal{O}(M \times N_S \times K)$, where $M$ is the (mean) number of MH steps used for each time step and smoothing particle. The advantage of the new procedure is that $M$ can often be much smaller $N_F$ and yet achieve the same error performance.

With the direct backward-resampling smoother, the only parameter to be chosen is $N_S$, the number of smoother particles. With the MCMC variant, we also have the freedom to set $M$, the number of MH steps in each Markov chain. This controls a trade-off between speed and particle diversity. If the chains are short, and the acceptance probabilities low, then this scheme will simply reproduce particles from the filter-smoother. However, when the weights of the conditional approximation, equation~\ref{eq:MCMC-BRS_weights} are similar, then an independent sample may be produced with only a very short chain.

@@

The method described in \cite{Fearnhead2010} uses a forward and a backward particle filter. To develop a a marginal smoothing approximation at time step $k$, particles are resampled from the forward filter at step $k-1$, and from the backward filter and step $k+1$. A new state for step $k$ is then sampled from a continuous proposal distribution conditional on the previous and next states, and an importance weight is assigned. It is undesirable to use importance sampling for generating particles from the joint sampling distribution, as the required resampling is liable to lead to a loss of particle diversity similar to that suffered by the filter-smoother.

The proposed new method proceeds in a similar manner to algorithm~\ref{alg:MCMC-BRS}. However, this time $x_{1:k-1}$, is resampled from the particle filter-smoother approximation from time step $k-1$. The current state, $x_k$, is then sampled afresh from a new proposal distribution. Thus the factorised proposal may be written as follows:

@@

The complete MCMC backward-sampling smoother is summarised in algorithm~\ref{alg:MCMC-BSS}.

\begin{algorithm}
  \begin{algorithmic}
  	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
		\FOR{$i = 1 \dots N_S$}
			\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
			\FOR{$k = K-1 \dots 1$}
				\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
				\FOR{$m = 1 \dots M$}
					\STATE Propose a new state hisory, $x_{1:k-1}^{(i)*} \sim \sum_j \tilde{v}_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1})$.
					\STATE Propose a new state, $x_{1:k}^{(i)*} \sim q(x_k|x_{k-1}^{*},\tilde{x}_{k+1},y_k)$.
					\STATE With probability $\alpha = \frac{ w_{k-1}^{*} \tilde{v}_{k-1}^{(m)} }{  w_{k-1}^{(m)} \tilde{v}_{k-1}^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) }$, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
				\ENDFOR
				\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{k}^{(i)(M)}$
			\ENDFOR
		\ENDFOR
  \end{algorithmic}
  \caption{MCMC backward-sampling smoother algorithm}
  \label{alg:MCMC-BSS}
\end{algorithm}

@@

\begin{itemize}
	\item The simple filter-smoother of \cite{Kitagawa1996}
	\item The direct backward-resampling smoother of \cite{Godsill2004}
	\item The new MCMC backward resampling smoother of section~\ref{sec:mcmc_smoother}, with several values of $M$
	\item The new MCMC backward sampling smoother of section~\ref{sec:new_state_smoother}, with several values of $M$
\end{itemize}

The algorithms were all used to generate XX particles from the joint smoothing distribution, using particle filters with XX particles. Where appropriate, proposal weights were chosen to be equal to filtering weights, i.e. $v_k^{(j)} = w_k^{(j)}$.

XX realisations were simulated from the model and the algorithms tested on each one with XX different random seeds.

Tables~\ref{} and~\ref{} show the root-mean-square errors (RMSEs) in position and velocity obtained by the different algorithms for two different settings of $\mathbf{Q}$ and $\mathbf{R}$, while tables~\ref{} and~\ref{} show the corresponding running times. The trade-off between speed and error is shown in figure~\ref{}. The mean number of particles over time is shown in figure~\ref{}.

@@

\section{Conclusion} \label{sec:conclusions}
The conclusion goes here.

>>>>

\title{Improved Particle Approximations \\ to the Joint Smoothing Distribution \\ Using Markov Chain Monte Carlo}

@@

\begin{abstract}
Particle filtering and smoothing algorithms approximate posterior state distributions with a set of samples drawn from those distributions. Conventionally, samples from the joint smoothing distribution are generated by sequentially resampling from the particle filter results. If the number of filtering particles is high, this process is limited by computational complexity. In addition, the support of the smoothing distribution is restricted to the values which appear in the filtering approximation. In this paper, a Markov chain Monte Carlo sampling procedure is used to improve the efficiency of the particle smoother. In addition, an algorithm for approximating the joint smoothing distribution without limited support is presented.
\end{abstract}

\begin{IEEEkeywords}
state space model, particle filter, smoothing, MCMC, Bayesian inference
\end{IEEEkeywords}

@@

Each particle is an entire realisation of the state process, with $K$ values corresponding to the state at each time step. Methods for generating these samples have been described in \cite{Kitagawa1996,Godsill2004,Briers2010}. In this paper new methods are presented for drawing samples from the joint smoothing distribution, using Markov chain Monte Carlo (MCMC). These novel procedures have computational advantages and greater flexibility over previous methods.

In section~\ref{sec:basics}, we briefly review the basic particle filter and existing joint smoother algorithms. The new MCMC-based smoothing methods are presented in sections~\ref{sec:mcmc_smoother} and~\ref{sec:new_state_smoother} with supporting simulations in section~\ref{sec:simulations}.

@@

The particle filter is summarised below.%in algorithm~\ref{alg:PF}.

%\begin{algorithm}
\begin{algorithmic}
	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
	\FOR{$k = 1 \dots K$}
 		\FOR{$i = 1 \dots N_F$}
 			\STATE Sample $\{ x_{k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 			\STATE Discard $x_{k-1}^{(i)}$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
\end{algorithmic}
%\caption{Particle filter algorithm}
%\label{alg:PF}
%\end{algorithm}

@@

The filter-smoother is summarised below.%in algorithm~\ref{alg:FS}.

%\begin{algorithm}
\begin{algorithmic}
 	\STATE Initialise particles from prior, $x_{0}^{(i)} \sim p(x_{0})$.
 	\FOR{$k = 1 \dots K$}
 		\FOR{$i = 1 \dots N_F$}
 			\STATE Sample $\{ x_{1:k-1}^{(i)}, x_k^{(i)} \} \sim \sum_j v_{k-1}^{(j)} q(x_k|x_{k-1}^{(j)})$.
 			\STATE Weight $w_{k}^{(i)} \propto \frac{ p(y_k|x_k^{(i)}) p(x_k^{(i)}|x_{k-1}^{(i)}) }{ q(x_{x}^{(i)}|x_{k-1}^{(i)}) } \times \frac{w_{k-1}^{(i)}}{v_{k-1}^{(i)} }$.
 		\ENDFOR
 	  \STATE Scale weights so that $\sum_i w_{k}^{(i)} = 1$.
 	\ENDFOR
 \end{algorithmic}
%\caption{Filter-smoother algorithm}
%\label{alg:FS}
%\end{algorithm}

An improved particle smoothing algorithm was presented in \cite{Godsill2004}. This works by constructing new particles from the smoothing distribution by resampling from the filtering approximations. The sampling procedure exploits the following factorisation of the joint distribution.

@@

Particles are generated by sampling backwards in time from the factors of this expansion. At the $k^{th}$ step, a sample $\tilde{x}_{k+1:K} \sim p(x_{k+1:K}|y_{1:K})$ will already have been drawn. By sampling $x_k \sim p(x_k|\tilde{x}_{k+1}, y_{1:K})$ and appending $x_k$ to $\tilde{x}_{k+1:K}$, the state sequence is sequentially extended back in time. The recursion is initialised with $x_K \sim p(x_K|y_{1:K})$, i.e. a sample from the final filtering distribution.

@@

The procedure may be repeated to produce as many samples from the joint smoothing distribution as required. This backward-resampling smoother is summarised below.% in algorithm~\ref{alg:DBRS}.

%\begin{algorithm}
\begin{algorithmic}
 	\STATE Run a particle filter to approximate $p(x_k|y_{1:k})$ for each $k$.
	\FOR{$i = 1 \dots N_S$}
		\STATE Sample $\tilde{x}_{K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{K}^{(j)}}(x_{K})$.
		\FOR{$k = K-1 \dots 1$}
			\FOR{$j = 1 \dots N_F$}
				\STATE Calculate weight $\tilde{w}_k^{(j)} = w_k^{(j)} p(\tilde{x}_{k+1}|x_k^{(j)})$.
			\ENDFOR
			\STATE Sample $\tilde{x}_{k}^{(i)} \sim \sum_j \tilde{w}_k^{(j)} \delta_{x_{k}^{(j)}}(x_{k})$.
		\ENDFOR
	\ENDFOR
\end{algorithmic}
%\caption{Direct backward-resampling smoother algorithm}
%\label{alg:DBRS}
%\end{algorithm}

A method for drawing samples from the joint smoothing distribution, $p(x_{1:K}|y_{1:K})$, is included in \cite{Briers2010}. The principal topic of this paper is a two-filter smoother for estimation of the marginal smoothing distributions, $p(x_{k}|y_{1:K})$, by combination of the results from one forward and one backward particle filter. Particles from the joint distribution may then be generated by resampling from either the forward or backward filters.% in the manner of algorithm~\ref{alg:DBRS}.

@@

%In section~\ref{sec:mcmc_smoother}, we introduce an MCMC approach for drawing samples from the joint smoothing distribution with an improved computational complexity over the direct backward-resampling smoother. In section~\ref{sec:new_state_smoother}, we combine this procedure with the new-proposal method used in \cite{Fearnhead2010} to generate joint smoothing distribution particles with full support.

@@

The bottleneck of the backward-resampling smoother is the calculation of the sampling weights (equation~\ref{eq:DBRS_weights}). For each smoother particle and for each time step, a weight must be calculated for each of the $N_F$ filter particles, in order to draw a sample from the conditional distribution $\hat{p}(x_k|\tilde{x}_{k+1}, y_{1:K})$. The innovation here is to use Markov chain Monte Carlo (MCMC) to produce these samples instead of sampling directly.

@@

The complete MCMC backward-resampling smoother is summarised below.% in algorithm~\ref{alg:MCMC-BRS}.

%\begin{algorithm}
\begin{algorithmic}
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
	\FOR{$i = 1 \dots N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1 \dots 1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1 \dots M$}
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim \sum_j \tilde{v}_k^{(j)} \delta_{x_{1:k}^{(j)}}(x_{1:k})$.
				\STATE With probability $\alpha = \frac{ w_k^{*} \tilde{v}_k^{(m)} }{  w_k^{(m)} \tilde{v}_k^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) }$, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
%\caption{MCMC backward-resampling smoother algorithm}
%\label{alg:MCMC-BRS}
%\end{algorithm}

@@

The computational complexity of the MCMC-based algorithm is $\mathcal{O}(M \times N_S \times K)$, where $M$ is the (mean) number of MH steps used for each time step and smoothing particle. The advantage of the new procedure is that $M$ can often be much smaller than $N_F$ and yet achieve almost equal error performance.

With the direct backward-resampling smoother, the only parameter to be chosen is $N_S$, the number of smoother particles. With the MCMC variant, we also have the freedom to set $M$, the number of MH steps in each Markov chain. This controls a trade-off between speed and particle diversity. If the chains are short, and the acceptance probabilities low, then this scheme will simply reproduce particles from the filter-smoother. However, when the weights of the conditional distribution approximation (equation~\ref{eq:MCMC-BRS_weights}) are similar, then an independent sample may be produced with only a very short chain.

@@

The method described in \cite{Fearnhead2010} uses a forward and a backward particle filter. To develop a marginal smoothing approximation at time step $k$, particles are resampled from the forward filter at step $k-1$, and from the backward filter and step $k+1$. A new state for step $k$ is then sampled from a continuous proposal distribution conditional on the previous and next states, and an importance weight is assigned. It is undesirable to use importance sampling for generating particles from the joint sampling distribution, as the required resampling is liable to lead to a loss of particle diversity similar to that suffered by the filter-smoother.

The proposed new method proceeds in a similar manner to the MCMC backward-resampling algorithm. However, this time $x_{1:k-1}$, is resampled from the particle filter-smoother approximation from time step $k-1$. The current state, $x_k$, is then sampled afresh from a new proposal distribution. Thus the factorised proposal may be written as follows:

@@

The complete MCMC backward-sampling smoother is summarised below.% in algorithm~\ref{alg:MCMC-BSS}.

%\begin{algorithm}
 \begin{algorithmic}
 	\STATE Run a particle filter to approximate $p(x_{1:k}|y_{1:k})$ for each $k$.
	\FOR{$i = 1 \dots N_S$}
		\STATE Sample $\tilde{x}_{1:K}^{(i)} \sim \sum_j w_K^{(j)} \delta_{x_{1:K}^{(j)}}(x_{1:K})$.
		\FOR{$k = K-1 \dots 1$}
			\STATE $x_{1:k}^{(i)(0)} \gets \tilde{x}_{1:k}^{(i)}$.
			\FOR{$m = 1 \dots M$}
				\STATE Propose a new state hisory, $x_{1:k-1}^{(i)*} \sim \sum_j \tilde{v}_{k-1}^{(j)} \delta_{x_{1:k-1}^{(j)}}(x_{1:k-1})$.
				\STATE Propose a new state, $x_{1:k}^{(i)*} \sim q(x_k|x_{k-1}^{*},\tilde{x}_{k+1},y_k)$.
				\STATE With probability $\alpha = \frac{ w_{k-1}^{*} \tilde{v}_{k-1}^{(m)} }{  w_{k-1}^{(m)} \tilde{v}_{k-1}^{*} } \times \frac{ p(\tilde{x}_{k+1}|x_k^{*}) p(x_k^{*}|x_{k-1}^{*}) p(y_k|x_k^{*}) q(x_k^{(m)}|x_{k-1}^{(m)},\tilde{x}_{k+1},y_k) }{ p(\tilde{x}_{k+1}|x_k^{(m)}) p(x_k^{(m)}|x_{k-1}^{(m)}) p(y_k|x_k^{(m)}) q(x_k^{*}|x_{k-1}^{*},\tilde{x}_{k+1},y_k) }$, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)*}$. Otherwise, $x_{1:k}^{(i)(m)} \gets x_{1:k}^{(i)(m-1)}$.
			\ENDFOR
			\STATE $\tilde{x}_{1:k}^{(i)} \gets x_{k}^{(i)(M)}$
		\ENDFOR
	\ENDFOR
\end{algorithmic}
%\caption{MCMC backward-sampling smoother algorithm}
%\label{alg:MCMC-BSS}
%\end{algorithm}

@@

\begin{itemize}
	\item The simple filter-smoother (FS) of \cite{Kitagawa1996}
	\item The direct backward-resampling smoother (DBRS) of \cite{Godsill2004}
	\item The new MCMC backward resampling smoother (MCMC-BRS) of section~\ref{sec:mcmc_smoother}, with several values of $M$
	\item The new MCMC backward sampling smoother (MCMC-BSS) of section~\ref{sec:new_state_smoother}, with several values of $M$
\end{itemize}

The algorithms were all used to generate 100 particles from the joint smoothing distribution, using particle filters with 100 particles. Where appropriate, proposal weights were chosen to be equal to filtering weights, i.e. $v_k^{(j)} = w_k^{(j)}$. The particle filter and MCMC-BSS both use ``optimal importance distributions'' (see e.g. \cite{Cappe2007}) with linearisations of the observation model where appropriate.

10 realisations of 500 time steps were simulated from the model and the algorithms tested on each one. Results are shown here for high and low observation noise cases. For case 1, the observation variances were set to $\sigma_B = \pi/720$ and $\sigma_R = 0.1$. For case 2, $\sigma_B = \pi/36$ and $\sigma_R = 100$. Other model settings are $\Delta t = 1$, $\sigma_P = 1$, and $x_0 = [-100, 50, 10, 0]^T$. Example tracjectories and observations from the two cases are shown in figure~\ref{fig:example_trajectories}.

\begin{figure}
\centerline{\subfloat[Case 1]{\includegraphics[width=0.45\columnwidth]{case1_trajectory-crop.pdf}%
\label{fig_first_case}}
\hfil
\subfloat[Case 2]{\includegraphics[width=0.45\columnwidth]{case2_trajectory-crop.pdf}%
\label{fig_second_case}}}
\caption{Example trajectory with observations from the two parameter cases.}
\label{fig:example_trajectories}
\end{figure}

Tables~\ref{tab:case1_performance} and~\ref{tab:case1_performance} show the root-mean-square errors (RMSEs) in position and velocity obtained by the different algorithms with the two parameter cases, along with running times (excluding the time for filtering) and the mean number of unique particles at each time instant. The trade-off between speed and error is shown in figures~\ref{fig:case1_rmse_vs_time} and~\ref{fig:case2_rmse_vs_time}.

\begin{table}%
\begin{tabular}{|c||c|c|c|c|}
\hline
Algorithm & \begin{minipage}[c]{1cm} Postion RMSE \end{minipage} & \begin{minipage}[c]{1cm}  Velocity RMSE \end{minipage} & \begin{minipage}[c]{1cm}  Running Time (s) \end{minipage} & \begin{minipage}[c]{1cm}  Unique Particles \end{minipage} \\
\hline
FS 						& 0.578 & 0.968 & 0 & 2.25 \\
DBRS					& 0.475 & 0.762 & 122 & 20.3 \\
\hline
MCMC-BRS (1)	& 0.509 & 0.823 & 3 & 13.7 \\
MCMC-BRS (3)	& 0.493 & 0.790 & 7 & 17.4 \\
MCMC-BRS (10)	& 0.482 & 0.768 & 21 & 19.6 \\
MCMC-BRS (30)	& 0.480 & 0.761 & 62 & 20.2 \\
MCMC-BRS (100)& 0.476 & 0.764 & 202 & 20.4 \\
\hline
MCMC-BSS (1)	& 0.473 & 0.758 & 14 & 44.3 \\
MCMC-BSS (3)	& 0.461 & 0.734 & 29 & 69.7 \\
MCMC-BSS (10)	& 0.451 & 0.720 & 82 & 89.8 \\
MCMC-BSS (30)	& 0.444 & 0.713 & 232 & 96.7 \\
MCMC-BSS (100)& 0.443 & 0.709 & 755 & 98.6 \\
\hline
\end{tabular}
\caption{Smoother performance measures for parameter case 1. All results are averaged over 10 runs, with 500 time steps in each run.}
\label{tab:case1_performance}
\end{table}

\begin{table}%
\begin{tabular}{|c||c|c|c|c|}
\hline
Algorithm & \begin{minipage}[c]{1cm} Postion RMSE \end{minipage} & \begin{minipage}[c]{1cm}  Velocity RMSE \end{minipage} & \begin{minipage}[c]{1cm}  Running Time (s) \end{minipage} & \begin{minipage}[c]{1cm}  Unique Particles \end{minipage} \\
\hline
FS 						& 0. & 0. & 0 &  \\
DBRS					& 0. & 0. &  &  \\
\hline
MCMC-BRS (1)	& 0. & 0. &  &  \\
MCMC-BRS (3)	& 0. & 0. &  &  \\
MCMC-BRS (10)	& 0. & 0. &  &  \\
MCMC-BRS (30)	& 0. & 0. &  &  \\
MCMC-BRS (100)& 0. & 0. &  &  \\
\hline
MCMC-BSS (1)	& 0. & 0. &  &  \\
MCMC-BSS (3)	& 0. & 0. &  &  \\
MCMC-BSS (10)	& 0. & 0. &  &  \\
MCMC-BSS (30)	& 0. & 0. &  &  \\
MCMC-BSS (100)& 0. & 0. &  &  \\
\hline
\end{tabular}
\caption{Smoother performance measures for parameter case 2. All results are averaged over 10 runs, with 500 time steps in each run.}
\label{tab:case2_performance}
\end{table}

\begin{figure}%
\includegraphics[width=\columnwidth]{case1_smoother_comparison_posRMSE_time-crop.pdf}%
\caption{RMSEs and running times of various smoothing algorithms}%
\label{fig:case1_rmse_vs_time}%
\end{figure}

\begin{figure}%
%\includegraphics[width=\columnwidth]{case2_smoother_comparison_posRMSE_time-crop.pdf}%
\caption{RMSEs and running times of various smoothing algorithms}%
\label{fig:case2_rmse_vs_time}%
\end{figure}

The MCMC-BRS smoother error approaches that of the DBRS as the number of MCMC steps, $M$, increases. However, for small $M$, the error is only very slightly larger than that of the DBRS for very low running times. For a particular application, it is possible to adjust $M$ to control the trade-off between error and speed.

The MCMC-BSS outperforms all the other smoothers, achieveing a lower RMSE even with $M=1$. As $M$ is increased further, the accuracy increases.

\section{Conclusion} \label{sec:conclusions}
Two new algorithms have been presented for drawing particles from the joint smoothing distribution using MCMC. The MCMC schemes have a degree of flexibility in the choice of chain length, which controls a trade-off between accuracy and running time.

The first algorithm resamples particles from the filtering distributions to produce particles from the smoothing distribution. The error performance of this scheme approaches that of the standard resampling scheme of \cite{Godsill2004} as the chain length increases. However, comparable error performance may be achieved with very short chains with a significant saving in processing time.

The second algorithm samples new values of the state from a proposal distribution. Although slower than the resampling version for a fixed length of Markov chain, the error performance is reduced relative to the other smoothers.

@@
